A family of accessibility measures derived from spatial interaction principles

### 2025-09-03 paper_plos/paper_plos_V0.qmd
final changes to R1

---
title: "A family of accessibility measures derived from spatial interaction principles"
#subtitle: "Submitted preprint for peer-review (May 2025)"
author:
  - name: Anastasia Soukhov
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Rafael H. M. Pereira
    email: rafael.pereira@ipea.gov.br
    affiliations:
        - name: Institute for Applied Economic Research - Ipea
          city: Brasília, Brazil
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliations:
        - name: University of Toronto Scarborough
          department: Department of Human Geography
          city: Toronto, Canada
    attributes:
        corresponding: true
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format:
  pdf:
    keep-tex: true
    documentclass: article
    header-includes: |
      \usepackage{float}
      \floatplacement{table}{H}
      \floatplacement{figure}{H}
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
#includes:
#      in_header: 'gt_packages.sty'
      #\usepackage{float}

@@

On the other hand, there is another subset of literature that cite both @hansen1959 and @wilson1971 that is accessibility-focused. We categorise their citation of @wilson1971 for three general reasons. Firstly, a group of these works cite @wilson1971 as attribution for using context-dependent travel cost functions [e.g., @weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining]. These works do not necessarily comment on spatial interaction explicitly. Secondly, another group of this accessibility-focused "both" citing literature _does_ associate spatial interaction as defined in @wilson1971 with accessibility's potential for spatial interaction more explicitly [e.g., @millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring]. We agree accessibility and spatial interaction are related topics: accessibility is an expression of its *potential* and the @wilson1971 paper briefly touches on the concept. However, in some of this literature, @hansen1959 and @wilson1971 are co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending], perhaps revealing the murkiness of the distinction between spatial interaction and the *potential for* spatial interaction in the literature. Thirdly, there is a group of accessibility-focused works that interpret @hansen1959's model as the singly- or doubly- constrained spatial interaction model's inverse balancing factor [e.g., @vickermanAccessibilityAttractionPotential1974]. This group cities the earlier spatial interaction works that make this connection and is especially prominent in the investigation of competitive accessibility topics [e.g., @karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling]. As outlined in preceding sections, we argue interpreting the singly- or doubly- constrained spatial interaction model's balancing factor as accessibility yields output values that are similarly plagued by interpretability issues.

Lastly, as an extension of the third reason within this group, only the works of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 use @wilson1971's balancing factors as a method for maintaining constraints on opportunities within the context of competitive accessibility. These works introduce the balancing factors as a mechanisms to ensure that opportunities at each destination are proportionally allocated to each zone (based on the proportion of population seeking opportunities and the relative travel impedance). This is to ensure that each zonal accessibility value is the sum of this proportional allocation from each destination, and that all zonal values ultimately sum to the total number of opportunities in the region. However, these balancing factors were deduced intuitively. These works do not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly constrained model (derived from entropy maximization). This equivalence is only discovered in hindsight, as will be demonstrated in this study. These two works also do not discuss other constrained cases that will be addressed in the next section.

@@

1. **The unconstrained case** with variants $V_i^0$ and $M_i^0$. This case is equivalent to @hansen1959's formulation (first variant) and @reilly1929methods's market potential formulation (second variant), whereby a balancing factor is neglected, so units of zonal values are in units of 'opportunities-by-travel-impedance-value' and 'population-by-travel-impedance-value'. In this case, no constraints are introduced to ensure that values of marginals are preserved.

2. **The total constrained case** with variant $V_i^T$ and variant $M_j^T$. The first variant $V_i^T$--the _total constrained accessible opportunities measure_--resembles @hansen1959's formulation but with an additional regional balancing factor $K^T$ term, defined as the ratio of the total number of opportunities in the region to the total sum of unconstrained accessibility values in the region. In this way, $K^T$ ensures that each zonal accessibility value is a proportion of the total opportunities in the region (i.e., the total marginal in @fig-analytical-device-conc-accessibility), requires no information about the population seeking opportunities, and is in units of 'opportunities' accessible. The second variant $M_j^T$ is the transpose of $i$ to $j$ of the first variant, effectively a constrained version of market potential and is referred to as _total constrained accessible population measure_. In this variant, each zonal accessibility value is a proportion of the total population in the region (maintained by $\hat K^T$), requires no information about the opportunities that are sought, and each zonal value is in units of 'population' accessible. 
  - When conceptualising _potential_ for spatial interaction--whether with opportunities or population--the total constrained case reflects the lowest level of restriction and hence maximum potential. The balancing factors $K^T$ and $\hat K^T$ only ensure that $V_{ij}^T$ and $M_{ji}^T$ values end up matching the _total sum_ of one of the single marginals, not the individual marginals themselves. For example, $K^T$ does not guarantee that all the opportunities accessibility at $V_{i=1}^T$ reflect a proportional allocation of the destination mass (i.e., _number_ of opportunities) at $j=1, 2, 3$. In some cases, an allocation of opportunities from a destination $j$ to $i=1$ could _exceed_ the number of opportunities at that $j$ (i.e., meaning that destination $j$ is very attractive and reachable for an $i$, relative other flows in the region). However, what $K^T$ _does_ ensure is that each $V_{ij}^T$ does not exceed the overall number of opportunities in the region--the total marginal. In other words, $V_{i=1}^T$ expresses the number of opportunities that origin $i=1$ could potentially interact with, as drawn from the entire regional opportunity mass, rather than constrained by individual destination totals. In this way, the _total constrained_ case reflects the maximum amount of _potential_ while still maintaining interpretable units.

3. **The singly constrained case**, with two variants $V_i^S$ (opportunity-constrained) and $M_j^S$ (population-constrained). The first variant is mathematically equivalent to the spatial availability measure [@soukhovIntroducingSpatialAvailability2023], and this variant's per capita form is equivalent to the popular 2SFCA measure [@luo2003; @shen1998]. Both variants can also be defined using either @hansen1959's or the market potential formulation but with balancing factor $B_j$ for $V_i^S$ or $A_i$ for $M_j^S$. These balancing factors ensure that the total marginal (green box in @fig-analytical-device-conc-accessibility) is maintained _as well as_ the values at one of the single marginals (destination-mass marginal for opportunity-constrained and origin-mass marginal for population constrained in @fig-analytical-device-conc-accessibility). In this way, the singly constrained case reflects a medium amount of potential, restricted by either values fitting the single destination-mass marginal or the single origin-mass marginal. 
  - The first variant $V_i^S$ includes a set of destination-side balancing factors $B_j$ which ensure opportunities (or destination masses) are allocated to each origin $i$ based on the population (i.e, origin mass) of the origin $i$ and travel impedance from that $j$ to all $i$s. Hence, $B_j$ ensures that each $V_i^S$ is the product sum of a _proportional share_ of opportunities allocated from each destination $j$ implicitly also being a share of the total opportunities in the region. Similar to the total constrained case, each $V_i^S$ and $V_ij^S$ accessibility value is expressed in units of 'opportunities' accessible. But unlike the total constrained case, the singly constrained case explicitly incorporates population by allocating a balanced share of a destination's total opportunities to populations at reachable origins.
  - The second variant $M_j^S$ includes a set of opportunity-side balancing factors $A_i$ which ensure population (or origin masses) are allocated to each destination $j$ based on the destination mass of the destination $j$ and all possible travel impedance from that $i$ to all $j$s. $A_i$ ensures similar, but transposed, so constraints are satisfied. Specifically, each $M_j^S$ represents both a share of the total regional population and the sum of balanced proportions of population from all origins, allocated to each destination based on the opportunities sought and travel impedance. Each zonal value is expressed in units of 'population' accessible.

@@

And lastly, in @eq-travel-behaviour-scenarios we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities. Accessibility will be calculated three times for each case, one assuming the most decay ($f_1(c_{ij})$), another assuming medium decay ($f_2(c_{ij})$), and a third assuming the least decaying travel behaviour ($f_3(c_{ij})$) for the entire region. A helpful analogy may be tying travel behaviour to the used mode's mobility potential, i.e., the most decaying travel behaviour ($f_1(c_{ij})$) would assume all travel in the region being done by foot, while calculating accessibility assuming the least decay ($f_3(c_{ij})$) would assume unfettered automobility. Or alternatively, these functions could represent travel behaviour on snowstorm-affected day ($f_1(c_{ij})$) for the entire region versus a clear, ideal travel day ($f_3(c_{ij})$). As an example of a discussion on how travel behaviour has been considered in accessibility measures cost of travel see @paez2012measuring.

@@

unc_acc |>
  gt(rowname_col = "oid") |>
  tab_stubhead(label = "Origin") |>
  tab_spanner(label = "{{V[_i^0]}}",
              columns = 2:4,
              level = 2) |>
  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}", 
              columns = 2,
              level = 1) |>
  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
              columns = 3,
              level = 1) |>
  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
              columns = 4,
              level = 1) |>
  fmt_number(decimals = 3) |>
  # cols_label(V_unc_i_1 = md("units: *f<sub>1</sub>(c<sub>ij</sub>)-weighted physicians*"),
  #            V_unc_i_2 = md("units: *f<sub>2</sub>(c<sub>ij</sub>)-weighted physicians*"),
  #            V_unc_i_3 = md("units: *f<sub>3</sub>(c<sub>ij</sub>)-weighted physicians*"),
  #            .fn = md)  |>
  cols_label(V_unc_i_1 = md("units: *physicians-minute^-3*"),
             V_unc_i_2 = md("units: *physicians-minute^-2*"),
             V_unc_i_3 = md("units: *physicians-minute^-0.1*"),
             .fn = md)  |>
  cols_align(align = "center",
             columns = 2:4) |>
  cols_width(
    stub() ~ px(60),
    everything() ~ px(180)
  ) |>
  grand_summary_rows(
    columns = c(V_unc_i_1, V_unc_i_2, V_unc_i_3),
    fns = list(label = "Sum", fn = "sum")) |> 
  tab_options(column_labels.font.size = "small",
              table.font.size = px(10)) |>
  as_latex()
```

@@

Which is also congruent with Wilson's framework as it comparable to the total flow spatial interaction model (e.g., Equation 2.11 in @cliff_evaluating_1974). Hence, rearranging the equation to have opportunities and the proportional constant distinctly represented, our total constrained accessibility model is:

@@

tot_acc |>
  select(oid, starts_with("V_")) |>
  gt(rowname_col = "oid") |>
  tab_stubhead(label = "Origin") |>
  tab_spanner(label = "{{V[_i^T]}}",
              columns = 2:4,
              level = 2) |>
  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
              columns = 2,
              level = 1) |>
  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
              columns = 3,
              level = 1) |>
  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
              columns = 4,
              level = 1) |>
  fmt_number(decimals = 3) |>
  cols_label(V_tot_i_1 = md("units: *physicians*"),
             V_tot_i_2 = md("units: *physicians*"),
             V_tot_i_3 = md("units: *physicians*"))  |>
  cols_align(align = "center",
             columns = 2:4) |>
  cols_width(
    stub() ~ px(60),
    everything() ~ px(180)
  ) |>
  tab_options(column_labels.font.size = "small",
              table.font.size = px(10)) |>
  grand_summary_rows(
    columns = c(V_tot_i_1, V_tot_i_2, V_tot_i_3),
    fns = list(label = "Sum", fn = "sum")) |> 
  as_latex()
```

@@

Using the total opportunity constrained formulation of accessibility offers a solution to the unit interpretability issue of @hansen1959's accessibility measure. Intuitively, the use of the constraint illustrates how the differences and ratios of values between zones and decay groups can be compared. This is true for other constrained cases of the family of accessibility measures.

@@

Another variant of the total constrained accessibility measure is the _total constrained accessible population measure_, which represents the transpose of $i$ to $j$ of the total constrained accessible opportunities measure. This variant, expressed in @eq-total-constrained-market, represents an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974, and which Reilly earlier referred to as 'potential trade territories' [@reilly1929methods]. This unconstrained form of market potential $M_j^0$ (@eq-unconstrained-market), effectively the $i$ to $j$ transpose of $V^0_{ij}$, has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road]. Put another way, market potential can also be thought of as a form of _passive accessibility_, indicating the number of people that can reach each destination. However, like $V_{ij}^0$, issues of unit interpretability arise in $M_j^0$'s unconstrained form. To address this, the constrained variant, the total constrained accessible population measure $M^T_{j}$, is introduced. To formulate this variant, the total balancing factor $K^T$ is applied to the mass of the *population* at $i$ ($W_i^{(1)}$) instead of the opportunities at $j$.

@@

tot_market |>
  select(did, starts_with("M_")) |>
  gt(rowname_col = "did") |>
  tab_stubhead(label = "Destination") |>
  tab_spanner(label = "{{M[_i^S]}}",
              columns = 2:4,
              level = 2) |>
  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
              columns = 2,
              level = 1) |>
  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
              columns = 3,
              level = 1) |>
  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
              columns = 4,
              level = 1) |>
  fmt_number(decimals = 3) |>
  cols_label(M_tot_i_1 = md("units: *population in 10,000s*"),
             M_tot_i_2 = md("units: *population in 10,000s*"),
             M_tot_i_3 = md("units: *population in 10,000s*"))  |>
  cols_align(align = "center",
             columns = 2:4) |>
  cols_width(
    stub() ~ px(90),
    everything() ~ px(150)
  ) |>
  grand_summary_rows(
    columns = c(M_tot_i_1, M_tot_i_2, M_tot_i_3),
    fns = list(label = "Sum", fn = "sum")) |> 
  tab_options(column_labels.font.size = "small",
              table.font.size = px(10)) |>
  as_latex()
```

@@

Similar to the total constrained accessibility measure, the singly constrained case includes a balancing factor that adjusts the unconstrained zonal accessibility values $V^0_i$ such that a _single_ constraint is satisfied. Two variants are defined: (a) the _singly constrained accessible opportunities_ case (an alternative formula to spatial availability in @soukhovIntroducingSpatialAvailability2023), and (b) the _singly constrained accessible population_ case, its transpose, or singly constrained market potential.

@@

The balancing factors $B_j$ for the $f_2(c_{ij})$ decay group for zones 1, 2 and 3 is `r B_j$B_j_2[1]`, `r B_j$B_j_2[2]` and `r B_j$B_j_2[3]`, and for $f_3(c_{ij})$ decay group is `r B_j$B_j_3[1]`, `r B_j$B_j_3[2]` and `r B_j$B_j_3[3]`. Using these these balancing constants, we can calculate the singly constrained opportunity accessibility:

@@

Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a *singly constrained* measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^{(1)}$ is now represented by $O_i$). In it's per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential been used to express "facility crowdedness" as in @wang_inverted_2018.

>>>>

---
title: "A family of accessibility measures derived from spatial interaction principles"
format:
  docx:
    toc: false
  plos-pdf:
    number-sections: false
    keep-tex: true
    journal:
      id: plosone
    header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
      - \floatplacement{figure}{H}
      - \usepackage{pdflscape}
      - \newcommand{\blandscape}{\begin{landscape}}
      - \newcommand{\elandscape}{\end{landscape}}
    #include-in-header:
    #  text: |
    #    % Remove comment for double spacing
    #    % \usepackage{setspace} 
    ##    % \doublespacing
    #    % \usepackage{pdflscape}
    #    % \newcommand{\blandscape}{\begin{landscape}}
    #    % \newcommand{\elandscape}{\end{landscape}}
author:
  - name: Anastasia Soukhov
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Rafael H. M. Pereira
    email: rafael.pereira@ipea.gov.br
    affiliations:
        - name: Institute for Applied Economic Research - Ipea
          department: Data Science Division
          city: Brasília, Brazil
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliations:
        - name: University of Toronto Scarborough
          department: Department of Human Geography
          city: Toronto, Canada
    attributes:
        corresponding: true
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada

author-notes:
  equal-contributor: These authors contributed equally to this work.
  deceased: Deceased
  group: Membership list can be found in the Acknowledgments sections

@@

On the other hand, there is another subset of literature that cite both Hansen [@hansen1959] and Wilson [@wilson1971] that is accessibility-focused. We categorise their citation of Wilson [@wilson1971] for three general reasons. Firstly, a group of these works cite Wilson [@wilson1971] as attribution for using context-dependent travel cost functions [e.g., @weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining]. These works do not necessarily comment on spatial interaction explicitly. Secondly, another group of this accessibility-focused "both" citing literature _does_ associate spatial interaction as defined in Wilson [@wilson1971] with accessibility's potential for spatial interaction more explicitly [e.g., @millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring]. We agree accessibility and spatial interaction are related topics: accessibility is an expression of its *potential* and the Wilson [@wilson1971] paper briefly touches on the concept. However, in some of this literature, Hansen [@hansen1959] and Wilson [@wilson1971] are co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending], perhaps revealing the murkiness of the distinction between spatial interaction and the *potential for* spatial interaction in the literature. Thirdly, there is a group of accessibility-focused works that interpret Hansen's [@hansen1959] model as the singly- or doubly- constrained spatial interaction model's inverse balancing factor [e.g., @vickermanAccessibilityAttractionPotential1974]. This group cities the earlier spatial interaction works that make this connection and is especially prominent in the investigation of competitive accessibility topics [e.g., @karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling]. As outlined in preceding sections, we argue interpreting the singly- or doubly- constrained spatial interaction model's balancing factor as accessibility yields output values that are similarly plagued by interpretability issues.

Lastly, as an extension of the third reason within this group, only the works of Soukhov et al. [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024] use Wilson's [@wilson1971] balancing factors as a method for maintaining constraints on opportunities within the context of competitive accessibility. These works introduce the balancing factors as a mechanisms to ensure that opportunities at each destination are proportionally allocated to each zone (based on the proportion of population seeking opportunities and the relative travel impedance). This is to ensure that each zonal accessibility value is the sum of this proportional allocation from each destination, and that all zonal values ultimately sum to the total number of opportunities in the region. However, these balancing factors were deduced intuitively. These works do not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly constrained model (derived from entropy maximization). This equivalence is only discovered in hindsight, as will be demonstrated in this study. These two works also do not discuss other constrained cases that will be addressed in the next section.

@@

1. **The unconstrained case** with variants $V_i^0$ and $M_i^0$. This case is equivalent to Hansen's [@hansen1959] formulation (first variant) and Reilly's [@reilly1929methods] market potential formulation (second variant), whereby a balancing factor is neglected, so units of zonal values are in units of 'opportunities-by-travel-impedance-value' and 'population-by-travel-impedance-value'. In this case, no constraints are introduced to ensure that values of marginals are preserved.

2. **The total constrained case** with variant $V_i^T$ and variant $M_j^T$. The first variant $V_i^T$--the _total constrained accessible opportunities measure_--resembles Hansen's [@hansen1959] formulation but with an additional regional balancing factor $K^T$ term, defined as the ratio of the total number of opportunities in the region to the total sum of unconstrained accessibility values in the region. In this way, $K^T$ ensures that each zonal accessibility value is a proportion of the total opportunities in the region (i.e., the total marginal in @fig-analytical-device-conc-accessibility), requires no information about the population seeking opportunities, and is in units of 'opportunities' accessible. The second variant $M_j^T$ is the transpose of $i$ to $j$ of the first variant, effectively a constrained version of market potential and is referred to as _total constrained accessible population measure_. In this variant, each zonal accessibility value is a proportion of the total population in the region (maintained by $\hat K^T$), requires no information about the opportunities that are sought, and each zonal value is in units of 'population' accessible. 
  - When conceptualising _potential_ for spatial interaction--whether with opportunities or population--the total constrained case reflects the lowest level of restriction and hence maximum potential. The balancing factors $K^T$ and $\hat K^T$ only ensure that $V_{ij}^T$ and $M_{ji}^T$ values end up matching the _total sum_ of one of the single marginals, not the individual marginals themselves. For example, $K^T$ does not guarantee that all the opportunities accessibility at $V_{i=1}^T$ reflect a proportional allocation of the destination mass (i.e., _number_ of opportunities) at $j=1, 2, 3$. In some cases, an allocation of opportunities from a destination $j$ to $i=1$ could _exceed_ the number of opportunities at that $j$ (i.e., meaning that destination $j$ is very attractive and reachable for an $i$, relative other flows in the region). However, what $K^T$ _does_ ensure is that each $V_{ij}^T$ does not exceed the overall number of opportunities in the region--the total marginal. In other words, $V_{i=1}^T$ expresses the number of opportunities that origin $i=1$ could potentially interact with, as drawn from the entire regional opportunity mass, rather than constrained by individual destination totals. In this way, the _total constrained_ case reflects the maximum amount of _potential_ while still maintaining interpretable units.

3. **The singly constrained case**, with two variants $V_i^S$ (opportunity-constrained) and $M_j^S$ (population-constrained). The first variant is mathematically equivalent to the spatial availability measure [@soukhovIntroducingSpatialAvailability2023], and this variant's per capita form is equivalent to the popular 2SFCA measure [@luo2003; @shen1998]. Both variants can also be defined using either Hansen's [@hansen1959] or the market potential formulation but with balancing factor $B_j$ for $V_i^S$ or $A_i$ for $M_j^S$. These balancing factors ensure that the total marginal (green box in @fig-analytical-device-conc-accessibility) is maintained _as well as_ the values at one of the single marginals (destination-mass marginal for opportunity-constrained and origin-mass marginal for population constrained in @fig-analytical-device-conc-accessibility). In this way, the singly constrained case reflects a medium amount of potential, restricted by either values fitting the single destination-mass marginal or the single origin-mass marginal. 
  - The first variant $V_i^S$ includes a set of destination-side balancing factors $B_j$ which ensure opportunities (or destination masses) are allocated to each origin $i$ based on the population (i.e, origin mass) of the origin $i$ and travel impedance from that $j$ to all $i$s. Hence, $B_j$ ensures that each $V_i^S$ is the product sum of a _proportional share_ of opportunities allocated from each destination $j$ implicitly also being a share of the total opportunities in the region. Similar to the total constrained case, each $V_i^S$ and $V_ij^S$ accessibility value is expressed in units of 'opportunities' accessible. But unlike the total constrained case, the singly constrained case explicitly incorporates population by allocating a balanced share of a destination's total opportunities to populations at reachable origins.
  - The second variant $M_j^S$ includes a set of opportunity-side balancing factors $A_i$ which ensure population (or origin masses) are allocated to each destination $j$ based on the destination mass of the destination $j$ and all possible travel impedance from that $i$ to all $j$s. $A_i$ ensures similar, but transposed, so constraints are satisfied. Specifically, each $M_j^S$ represents both a share of the total regional population and the sum of balanced proportions of population from all origins, allocated to each destination based on the opportunities sought and travel impedance. Each zonal value is expressed in units of 'population' accessible.

@@

And lastly, in @eq-travel-behaviour-scenarios we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities. Accessibility will be calculated three times for each case, one assuming the most decay ($f_1(c_{ij})$), another assuming medium decay ($f_2(c_{ij})$), and a third assuming the least decaying travel behaviour ($f_3(c_{ij})$) for the entire region. A helpful analogy may be tying travel behaviour to the used mode's mobility potential, i.e., the most decaying travel behaviour ($f_1(c_{ij})$) would assume all travel in the region being done by foot, while calculating accessibility assuming the least decay ($f_3(c_{ij})$) would assume unfettered automobility. Or alternatively, these functions could represent travel behaviour on snowstorm-affected day ($f_1(c_{ij})$) for the entire region versus a clear, ideal travel day ($f_3(c_{ij})$). As an example of a discussion on how travel behaviour has been considered in accessibility measures cost of travel see Páez et al. [@paez2012measuring].

@@

unc_acc |>
  gt(rowname_col = "oid") |>
  tab_stubhead(label = "Origin") |>
  tab_spanner(label = "{{V[_i^0]}}",
              columns = 2:4,
              level = 2) |>
  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}", 
              columns = 2,
              level = 1) |>
  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
              columns = 3,
              level = 1) |>
  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
              columns = 4,
              level = 1) |>
  fmt_number(decimals = 3) |>
  # cols_label(V_unc_i_1 = md("units: *f<sub>1</sub>(c<sub>ij</sub>)-weighted physicians*"),
  #            V_unc_i_2 = md("units: *f<sub>2</sub>(c<sub>ij</sub>)-weighted physicians*"),
  #            V_unc_i_3 = md("units: *f<sub>3</sub>(c<sub>ij</sub>)-weighted physicians*"),
  #            .fn = md)  |>
  cols_label(V_unc_i_1 = md("units: *physicians-minute^-3*"),
             V_unc_i_2 = md("units: *physicians-minute^-2*"),
             V_unc_i_3 = md("units: *physicians-minute^-0.1*"),
             .fn = md)  |>
  cols_align(align = "center",
             columns = 2:4) |>
  #cols_width(
  #  stub() ~ px(60),
  #  everything() ~ px(180)
  #) |>
  grand_summary_rows(
    columns = c(V_unc_i_1, V_unc_i_2, V_unc_i_3),
    fns = list(label = "Sum", fn = "sum")) |> 
  tab_options(column_labels.font.size = "small",
              table.font.size = px(8)) |>
  as_latex()
```

@@

Which is also congruent with Wilson's framework as it comparable to the total flow spatial interaction model (e.g., Equation 2.11 in [@cliff_evaluating_1974]). Hence, rearranging the equation to have opportunities and the proportional constant distinctly represented, our total constrained accessibility model is:

@@

tot_acc |>
  select(oid, starts_with("V_")) |>
  gt(rowname_col = "oid") |>
  tab_stubhead(label = "Origin") |>
  tab_spanner(label = "{{V[_i^T]}}",
              columns = 2:4,
              level = 2) |>
  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
              columns = 2,
              level = 1) |>
  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
              columns = 3,
              level = 1) |>
  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
              columns = 4,
              level = 1) |>
  fmt_number(decimals = 3) |>
  cols_label(V_tot_i_1 = md("units: *physicians*"),
             V_tot_i_2 = md("units: *physicians*"),
             V_tot_i_3 = md("units: *physicians*"))  |>
  cols_align(align = "center",
             columns = 2:4) |>
  #cols_width(
  #  stub() ~ px(60),
  #  everything() ~ px(180)
  #) |>
  tab_options(column_labels.font.size = "small",
              table.font.size = px(8)) |>
  grand_summary_rows(
    columns = c(V_tot_i_1, V_tot_i_2, V_tot_i_3),
    fns = list(label = "Sum", fn = "sum")) |> 
  as_latex()
```

@@

Using the total opportunity constrained formulation of accessibility offers a solution to the unit interpretability issue of Hansen's [@hansen1959] accessibility measure. Intuitively, the use of the constraint illustrates how the differences and ratios of values between zones and decay groups can be compared. This is true for other constrained cases of the family of accessibility measures.

@@

Another variant of the total constrained accessibility measure is the _total constrained accessible population measure_, which represents the transpose of $i$ to $j$ of the total constrained accessible opportunities measure. This variant, expressed in @eq-total-constrained-market, represents an expression of the concept of market potential (i.e., potential users) as proposed in Harris [@harris_market_1954] and Vickerman [@vickermanAccessibilityAttractionPotential1974], and which Reilly earlier referred to as 'potential trade territories' [@reilly1929methods]. This unconstrained form of market potential $M_j^0$ (@eq-unconstrained-market), effectively the $i$ to $j$ transpose of $V^0_{ij}$, has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road]. Put another way, market potential can also be thought of as a form of _passive accessibility_, indicating the number of people that can reach each destination. However, like $V_{ij}^0$, issues of unit interpretability arise in $M_j^0$'s unconstrained form. To address this, the constrained variant, the total constrained accessible population measure $M^T_{j}$, is introduced. To formulate this variant, the total balancing factor $K^T$ is applied to the mass of the *population* at $i$ ($W_i^{(1)}$) instead of the opportunities at $j$.

@@

tot_market |>
  select(did, starts_with("M_")) |>
  gt(rowname_col = "did") |>
  tab_stubhead(label = "Destination") |>
  tab_spanner(label = "{{M[_i^S]}}",
              columns = 2:4,
              level = 2) |>
  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
              columns = 2,
              level = 1) |>
  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
              columns = 3,
              level = 1) |>
  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
              columns = 4,
              level = 1) |>
  fmt_number(decimals = 3) |>
  cols_label(M_tot_i_1 = md("units: *population in 10,000s*"),
             M_tot_i_2 = md("units: *population in 10,000s*"),
             M_tot_i_3 = md("units: *population in 10,000s*"))  |>
  cols_align(align = "center",
             columns = 2:4) |>
  #cols_width(
  #  stub() ~ px(90),
  #  everything() ~ px(150)
  #) |>
  grand_summary_rows(
    columns = c(M_tot_i_1, M_tot_i_2, M_tot_i_3),
    fns = list(label = "Sum", fn = "sum")) |> 
  tab_options(column_labels.font.size = "small",
              table.font.size = px(8)) |>
  as_latex()
```

@@

Similar to the total constrained accessibility measure, the singly constrained case includes a balancing factor that adjusts the unconstrained zonal accessibility values $V^0_i$ such that a _single_ constraint is satisfied. Two variants are defined: (a) the _singly constrained accessible opportunities_ case (an alternative formula to spatial availability in Soukhov et al. [@soukhovIntroducingSpatialAvailability2023]), and (b) the _singly constrained accessible population_ case, its transpose, or singly constrained market potential.

@@

The balancing factors $B_j$ for the $f_2(c_{ij})$ decay group for zones 1, 2 and 3 is `r B_j$B_j_2[1]`, `r B_j$B_j_2[2]` and `r B_j$B_j_2[3]`, and for $f_3(c_{ij})$ decay group is `r B_j$B_j_3[1]`, `r B_j$B_j_3[2]` and `r B_j$B_j_3[3]`. Using these these balancing constants, we can calculate the singly constrained opportunity accessibility (@tbl-simple-example-attraction-constrained-accessibility).

@@

Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a *singly constrained* measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^{(1)}$ is now represented by $O_i$). In it's per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential been used to express "facility crowdedness" as in Wang [@wang_inverted_2018].
